---
output:
  xaringan::moon_reader:
    css: "my-theme.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)
library(huxtable)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center")
library(MARSS)
```


class: center, middle

# Size and abundance of resident salmon in the Salish Sea: Insights from a culturally unique recreational fishery

## Mark Scheuerell

USGS Washington Cooperative Fish and Wildlife Research Unit<br>School of Aquatic and Fishery Sciences<br>University of Washington<br>Seattle, WA

.futnote.purple-text[scheuerl@uw.edu]

.citation.blue-text[@mark_scheuerell]

---

# Questions

.blue-text[
## Has size or catch rate changed over the decades?

## Do any changes mirror patterns in size or abundance of Puget Sound Chinook salmon as a whole?
]

---

class: center, middle, inverse

# Temporal trends in derby harvest

---

# Catch per unit effort

.blue-text[
## Things we know:

* ### Total number of days the derby was open

* ### Total number of participants

* ### Total number of fish caught
]

---

# Catch per unit effort

.blue-text[
## Things we know:

* ### Total number of days the derby was open

* ### Total number of participants

* ### Total number of fish caught
]

.red-text[
## Things we don't know:

* ### Total number of days that *each* angler fished
]

---

# Catch per unit effort

.blue-text[
## Thus, we assumed that each angler fished every day the derby was open
]

$$
\Large{
\text{CPUE} = \frac{\text{fish caught}}{\text{days} \times \text{anglers}}
}
$$

---

# Time series of CPUE

```{r plot_cpue, fig.dim=c(9, 6)}
## set data dir
datadir <- here::here("data")
## import raw Tengu data
tengu_data <- readr::read_csv(file.path(datadir, "tengu_derby_data.csv"))
## import raw WDFW data
wdfw_data <- readr::read_csv(file.path(datadir, "wdfw_data.csv"))

## CPUE
cpue <- (tengu_data$total_catch / (tengu_data$days * tengu_data$members))[-c(1:5)]

## time series plot of CPUE
par(mai = c(0.9, 1.1, 0.1, 0.1), cex.axis = 1.5, cex.lab = 1.8)
plot.ts(ts(cpue, start = 1950), lwd = 2, col = "dodgerblue",
        las = 1, log = "y", # xlim = c(1950,2020),
        ylab = "", xlab = "Year", yaxt = "n")
mtext("CPUE", side = 2, line = 3.5, cex = 1.6)
axis(2, at = c(0.01, 0.02, 0.05, 0.1), las = 1)
```


---

# Possible state models

## Random walk

$\Large{x_t = x_{t-1} + w_t}$



---

# Possible state models

## Random walk

$\Large{x_t = x_{t-1} + w_t}$

## Biased random walk

$\Large{x_t = x_{t-1} + u + w_t}$


---

# Observation model

### Imperfect accounting of effort means we are unsure of the true CPUE

$\Large{y_t = x_t + v_t}$


---

# Model fitting

### MARSS package for R (Holmes, Ward & Scheuerell 2020)

https://CRAN.R-project.org/package=MARSS

---

# CPUE has decreased by ~2% per year

```{r fit_cpue, cache = TRUE, results = 'hide'}
## model setup
mod_list <- list(
  B = matrix(1),
  U = matrix("u"),
  Q = matrix("q"),
  Z = matrix(1),
  A = matrix(0),
  R = matrix("r")
)

## log the response
l_cpue <- matrix(log(cpue), nrow = 1)

## fit biased RW
cpue_brw <- MARSS(l_cpue, model = mod_list)
```

```{r plot_cpue_fit, fig.dim=c(9, 6)}
## time series plot of CPUE
par(mai = c(0.9, 1.1, 0.1, 0.1), cex.axis = 1.5, cex.lab = 1.8)
plot.ts(ts(cpue, start = 1950), lwd = 2, col = "dodgerblue",
        las = 1, log = "y", xlim = c(1950,2020),
        ylab = "", xlab = "Year", yaxt = "n")
mtext("CPUE", side = 2, line = 3.5, cex = 1.6)
axis(2, at = c(0.01, 0.02, 0.05, 0.1), las = 1)
lines(ts(t(exp(cpue_brw$states)), start = 1950))
```

---

class: center, middle, inverse

# Temporal trends in derby size

---

# 3 metrics of size

.blue-text[
## 1) The weight of each of the top five Chinook salmon

## 2) The number over 10 pounds

## 3) The number over 5 pounds 
]

---

# Three metrics of fish size

.gray-text[
## 1) The weight of each of the top five Chinook salmon
]

.blue-text[
## 2) The number over 10 pounds

## 3) The number over 5 pounds 
]

---

# Proportions of large fish

---

# Time series of proportions

---

# Model forms

---

# Data support

---

# Model fit to data

---

class: middle, center, inverse

# Comparisons of derby and survey data

---

# WDFW surveys

## Losee et al (2019)

---

# Time series of CPUE

---

# Model forms

---

# Data support

---

# Model fit to data

---

# Time series of size

---

# Model forms

---

# Data support

---

# Model fit to data

---

# Conclusions

