---
output:
  xaringan::moon_reader:
    css: "my-theme.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)
library(huxtable)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center", dpi = 300)
library(MARSS)
library(dplyr)
```


class: center, middle

# Size and abundance of resident Chinook salmon in the Salish Sea: Insights from a culturally unique recreational fishery

## Mark Scheuerell

USGS Washington Cooperative Fish and Wildlife Research Unit<br>School of Aquatic and Fishery Sciences<br>University of Washington<br>Seattle, WA

.futnote.purple-text[`r icon::fa("envelope")` scheuerl@uw.edu]

.citation.blue-text[`r icon::fa("twitter")` @mark_scheuerell]

---

# Disclaimer

.red-text[
## I take full responsibility for this analysis
]

---

# Questions

.blue-text[
## Has size or catch rate changed over the decades?

## Do any changes mirror patterns in size or abundance of Puget Sound Chinook salmon as a whole?
]

---

class: center, middle, inverse

# Temporal trends in derby harvest

---

# Catch per unit effort

.blue-text[
## Things we know:

* ### Total number of days the derby was open

* ### Total number of participants

* ### Total number of fish caught
]

---

# Catch per unit effort

.blue-text[
## Things we know:

* ### Total number of days the derby was open

* ### Total number of participants

* ### Total number of fish caught
]

.red-text[
## Things we don't know:

* ### Total number of days that *each* angler fished
]

---

# Catch per unit effort

.blue-text[
## Thus, we assumed that each angler fished every day the derby was open
]

$$
\Large{
\text{CPUE} = \frac{\text{fish caught}}{\text{days} \times \text{anglers}}
}
$$

---

# Time series of CPUE

```{r plot_cpue, fig.dim = c(8, 5), }
## set data dir
datadir <- here::here("data")
## import raw Tengu data
tengu_data <- readr::read_csv(file.path(datadir, "tengu_derby_data.csv"))
## import raw WDFW data
wdfw_data <- readr::read_csv(file.path(datadir, "wdfw_data.csv"))

## CPUE
cpue <- (tengu_data$total_catch / (tengu_data$days * tengu_data$members))[-c(1:5)]

## time series plot of CPUE
par(mai = c(1.1, 1.3, 0.3, 0.3), cex.axis = 1.3, cex.lab = 1.6)
plot.ts(ts(cpue, start = 1950), lwd = 2, col = "dodgerblue",
        las = 1, log = "y", # xlim = c(1950,2020),
        ylab = "", xlab = "Year", yaxt = "n")
mtext("CPUE", side = 2, line = 3.5, cex = 1.8)
axis(2, at = c(0.01, 0.02, 0.05, 0.1), las = 1)
```


---

# Possible state models

## Random walk

$\Large{x_t = x_{t-1} + w_t}$

<br>

where $x_t$ is the log(CPUE) at time $t$

---

# Possible state models

## Random walk

$\Large{x_t = x_{t-1} + w_t}$

## Biased random walk

$\Large{x_t = x_{t-1} + u + w_t}$


---

# Observation model

.red-text[
### Imperfect accounting of effort means we are unsure of the true CPUE
]

$\Large{y_t = x_t + v_t}$


---

# Full state-space model

$\Large{x_t = x_{t-1} + u + w_t}$

$\Large{y_t = x_t + v_t}$


---

# Model fitting

### MARSS package for R (Holmes, Ward & Scheuerell 2020)

https://CRAN.R-project.org/package=MARSS

---

# CPUE has decreased by ~2% per year

```{r fit_cpue, cache = TRUE, results = 'hide'}
## model setup
mod_list <- list(
  B = matrix(1),
  U = matrix("u"),
  Q = matrix("q"),
  Z = matrix(1),
  A = matrix(0),
  R = matrix("r")
)

## log the response
l_cpue <- matrix(log(cpue), nrow = 1)

## fit biased RW
cpue_brw <- MARSS(l_cpue, model = mod_list)
```

```{r plot_cpue_fit, fig.dim = c(8, 5)}
se <- sqrt(coef(cpue_brw)$R)

## time series plot of CPUE
par(mai = c(1.1, 1.3, 0.3, 0.3), cex.axis = 1.3, cex.lab = 1.6)
plot.ts(ts(cpue, start = 1950), lwd = 2, col = "dodgerblue",
        las = 1, log = "y", xlim = c(1950,2020),
        ylab = "", xlab = "Year", yaxt = "n")
mtext("CPUE", side = 2, line = 3.5, cex = 1.8)
axis(2, at = c(0.01, 0.02, 0.05, 0.1), las = 1)
lines(ts(t(exp(cpue_brw$states)), start = 1950), lwd = 1)
lines(ts(exp(as.vector(cpue_brw$states) + 2*se), start = 1950),
      col = "darkgray")
lines(ts(exp(as.vector(cpue_brw$states) - 2*se), start = 1950),
      col = "darkgray")
```

---

class: center, middle, inverse

# Temporal trends in derby size

---

# Three metrics of fish size

.blue-text[
## 1) The weight of each of the top five Chinook salmon*

## 2) The number over 10 pounds

## 3) The number over 5 pounds 
]

.citation[*In _most_ years]

---

# Three metrics of fish size

.gray-text[
## 1) The weight of each of the top five Chinook salmon
]

.blue-text[
## 2) The number over 10 pounds

## 3) The number over 5 pounds 
]

---

# Proportion of fish over 10 pounds

```{r plot_prop_over_10, fig.dim = c(8, 5)}
## proportion of fish >10 lbs
p10 <- (tengu_data$n_over_10 / tengu_data$total_catch)[-c(1:5)]
## screen for p = 0 & change to p = 0.001
p10[p10 == 0] <- 0.001

## ts of sizes
size_10 <- ts(p10, start = 1950)

## time series plot of fish size metrics
par(mai = c(1.1, 1.3, 0.3, 0.3), cex.axis = 1.3, cex.lab = 1.6)
plot.ts(size_10, lwd = 2, col = "dodgerblue",
        main = "", las = 1,
        ylab = c("", "", ""), xlab = "Year")
mtext("Proportion", side = 2, line = 3.5, cex = 1.8)
```

---

# Proportion of fish over 5 pounds

```{r plot_prop_over_5, fig.dim = c(8, 5)}
## proportion of fish >5 lbs
p5 <- (tengu_data$n_over_5 / tengu_data$total_catch)[-c(1:5)]
## screen for p = 1 & change to p = 0.999
p5[p5 == 1] <- 0.999
## ts of sizes
size_5 <- ts(p5, start = 1950)

## time series plot of fish size metrics
par(mai = c(1.1, 1.3, 0.3, 0.3), cex.axis = 1.3, cex.lab = 1.6)
plot.ts(size_5, lwd = 2, col = "dodgerblue",
        main = "", las = 1,
        ylab = c("", "", ""), xlab = "Year")
mtext("Proportion", side = 2, line = 3.5, cex = 1.8)
```

---

# Possible state models

## Random walk

$\Large{x_t = x_{t-1} + w_t}$

## Biased random walk

$\Large{x_t = x_{t-1} + u + w_t}$

<br>

where $x_t = \text{logit}(p_t)$ 

---

# Observation model

.red-text[
### Imperfect sampling means we are unsure of the true size
]

$\Large{y_t = x_t + v_t}$


---

```{r fit_tengu_sizes, cache = TRUE, results = "hide"}
## response for fish >10 lbs 
l_size_10 <- matrix(qlogis(p10), nrow = 1)

## response for fish >5 lbs
l_size_5 <- matrix(qlogis(p5), nrow = 1)

## model setup
mod_list <- list(
  B = matrix(1),
  U = matrix("u"),
  Q = matrix("q"),
  Z = matrix(1),
  A = matrix(0),
  R = matrix("r")
)

## over 10
## fit model with bias (Eqn 1)
size_brw_10 <- MARSS(l_size_10, model = mod_list)

## over 5
## fit model with bias (Eqn 1)
size_brw_5 <- MARSS(l_size_5, model = mod_list)

## set bias to 0
mod_list$U <- matrix(0)

## over 10
## fit model without bias (Eqn 2)
size_rw_10 <- MARSS(l_size_10, model = mod_list)

## over 5
## fit model without bias (Eqn 2)
size_rw_5 <- MARSS(l_size_5, model = mod_list)
```

# Proportion of fish over 10 pounds

```{r plot_prop_over_10_fit, fig.dim = c(8, 5)}
## time series plot of fish size metrics
par(mai = c(1.1, 1.3, 0.3, 0.3), cex.axis = 1.3, cex.lab = 1.6)
plot.ts(size_10, lwd = 2, col = "dodgerblue",
        main = "", las = 1,
        ylab = c("", "", ""), xlab = "Year")
mtext("Proportion", side = 2, line = 3.5, cex = 1.8)
lines(ts(t(plogis(size_rw_10$states)), start = 1950), lwd = 2)
lines(ts(t(plogis(size_rw_10$states + 2*size_rw_10$states.se)), start = 1950),
      col = "darkgray", lwd = 1)
lines(ts(t(plogis(size_rw_10$states - 2*size_rw_10$states.se)), start = 1950),
      col = "darkgray", lwd = 1)
```

---

# Proportion of fish over 5 pounds

```{r plot_prop_over_5_fit, fig.dim = c(8, 5)}
## time series plot of fish size metrics
par(mai = c(1.1, 1.3, 0.3, 0.3), cex.axis = 1.3, cex.lab = 1.6)
plot.ts(size_5, lwd = 2, col = "dodgerblue",
        main = "", las = 1,
        ylab = c("", "", ""), xlab = "Year")
mtext("Proportion", side = 2, line = 3.5, cex = 1.8)
lines(ts(t(plogis(size_rw_5$states)), start = 1950), lwd = 2)
lines(ts(t(plogis(size_rw_5$states + 2*size_rw_10$states.se)), start = 1950),
      col = "darkgray", lwd = 1)
lines(ts(t(plogis(size_rw_5$states - 2*size_rw_10$states.se)), start = 1950),
      col = "darkgray", lwd = 1)
```

---

class: middle, center, inverse

# Comparison of derby & WDFW catch data

---

# WDFW abundance estimates

```{r both_cpue, fig.dim = c(8, 5)}
## select common data
cpue_t <- seq(1950, 2019)
wdfw_t <- seq(1968, 2015)
yy <- cbind(tengu = cpue[cpue_t %in% wdfw_t],
            wdfw = wdfw_data$total)

## standardize the mean and variance
yy <- scale(log(yy))

par(mai = c(1.1, 1.3, 0.3, 0.3), cex.axis = 1.3, cex.lab = 1.6)
plot.ts(ts(wdfw_data$total/1000, start = 1968), lwd = 2, col = "indianred",
        main = "", las = 1,
        ylab = c("", "", ""), xlab = "Year")
mtext("Abundance (1000s)", side = 2, line = 3.5, cex = 1.8)
```

.citation.gray-text[Losee et al (2019) _Fish & Fisheries_]

---

# Time series of both abundance indices

```{r plot_wdfw_cpue, fig.dim = c(8, 5)}
par(mai = c(1.1, 1.3, 0.3, 0.3), cex.axis = 1.3, cex.lab = 1.6)
matplot(yy, type = "l", lty = "solid", lwd = 2, col = c("dodgerblue", "indianred"), las = 1,
        ylab = "Standardized abundance", xlab = "Year", xaxt = "n")
text(6, 1.4, "Tengu", pos = 3, cex = 1.5, col = "dodgerblue")
text(6, -1.7, "WDFW", pos = 1, cex = 1.5, col = "indianred")
axis(1, at = seq(1, 41, 10), labels = seq(1970, 2010, 10))
```

---

# Model forms

.blue-text[
## One common state
]

$$\begin{bmatrix} y_\text{Tengu} \\ y_\text{WDFW} \end{bmatrix}_t = \begin{bmatrix} 1 \\ 1 \end{bmatrix}
x_t + \begin{bmatrix} a_\text{Tengu} \\ a_\text{WDFW} \end{bmatrix} + \begin{bmatrix} v_\text{Tengu} \\
v_\text{WDFW} \end{bmatrix}_t$$

$$x_t = x_{t-1} + u + w_t$$

---

# Model forms

.blue-text[
## Two unique states
]

$$\begin{bmatrix} y_\text{Tengu} \\ y_\text{WDFW} \end{bmatrix}_t = \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix}
\begin{bmatrix} x_\text{Tengu} \\ x_\text{WDFW} \end{bmatrix}_t + \begin{bmatrix} a_\text{Tengu} \\ a_\text{WDFW} \end{bmatrix} + \begin{bmatrix} v_\text{Tengu} \\ v_\text{WDFW} \end{bmatrix}_t$$

$$\begin{bmatrix} x_\text{Tengu} \\ x_\text{WDFW} \end{bmatrix}_t = \begin{bmatrix} x_\text{Tengu} \\ x_\text{WDFW} \end{bmatrix}_{t-1} + \begin{bmatrix} u_\text{Tengu} \\ u_\text{WDFW} \end{bmatrix} + \begin{bmatrix} w_\text{Tengu} \\ w_\text{WDFW} \end{bmatrix}_t$$

---

class: middle, center

# The data support .green-text[two unique states] with a .green-text[common decline] of ~2.5% per year

---

# Model fits and data

```{r ssm_two_cpue, cache = TRUE, results = "hide"}
## model defn for Eqns 8 & 9
mod_list <- list(
  B = diag(2),
  # U = matrix(c("tengu", "wdfw"), nrow = 2, ncol = 1),
  U = matrix(c("bias", "bias"), nrow = 2, ncol = 1),
  # Q = matrix(list("T", 0, 0, "W"), 2, 2),
  Q = matrix(list("q", 0, 0, "q"), 2, 2),
  Z = diag(2),
  A = matrix(list(0, 0), nrow = 2, ncol = 1),
  # R = matrix(list("T", 0, 0, "W"), 2, 2)
  R = matrix(list("r", 0, 0, "r"), 2, 2)
)

## fit Eqns 8 & 9
cpue_both_2 <- MARSS(t(yy), model = mod_list, control = list(maxit = 1000))
```

```{r plot_ssm_two_cpue, fig.dim = c(8, 5), echo = FALSE}
par(mai = c(1.1, 1.3, 0.3, 0.3), cex.axis = 1.3, cex.lab = 1.6)
matplot(yy, type = "l", lty = "solid", lwd = 1, col = c("dodgerblue", "indianred"), las = 1,
        ylab = "Standardized abundance", xlab = "Year", xaxt = "n")
text(6, 1.4, "Tengu", pos = 3, cex = 1.5, col = "dodgerblue")
text(6, -1.7, "WDFW", pos = 1, cex = 1.5, col = "indianred")
matlines(t(cpue_both_2$states), type = "l", lty = "dashed", lwd = 2, col = c("dodgerblue", "indianred"))
axis(1, at = seq(1, 41, 10), labels = seq(1970, 2010, 10))
```


---

class: middle, center, inverse

# Comparison of derby & WDFW size data

---

# Three metrics of fish size

.blue-text[
## 1) The weight of each of the top five Chinook salmon*

## 2) The number over 10 pounds

## 3) The number over 5 pounds 
]

.citation[*In _most_ years]

---

# Three metrics of fish size

.blue-text[
## 1) The mean weight of the _top two_ Chinook salmon
]

.gray-text[
## 2) The number over 10 pounds

## 3) The number over 5 pounds 
]

---

# Time series of size

```{r plot_both_sizes, fig.dim = c(8, 5), echo = FALSE}
## total mass of 5 largest fish
mean_of_top_2 <- apply(tengu_data[,c(9:10)], 1, mean, na.rm = TRUE)
mean_of_top_2[is.nan(mean_of_top_2)] <- NA

## select common data
yy <- cbind(tengu = mean_of_top_2[tengu_data$year >= 1970 & tengu_data$year <= 2014],
            wdfw = wdfw_data$size[wdfw_data$year >= 1970 & wdfw_data$year <= 2014])

par(mai = c(1.1, 1.3, 0.3, 0.3), cex.axis = 1.3, cex.lab = 1.6)
matplot(yy, type = "l", lty = "solid", lwd = 2, col = c("dodgerblue", "indianred"), las = 1,
        ylab = "Mean size (kg)", xlab = "Year", xaxt = "n")
text(17, 4, "Tengu", pos = 1, cex = 1.5, col = "dodgerblue")
text(17, 8.2, "WDFW", pos = 3, cex = 1.5, col = "indianred")
axis(1, at = seq(1, 41, 10), labels = seq(1970, 2010, 10))
```

```{r ssm_two_size, cache = TRUE, results = "hide"}
## model defn
mod_list <- list(
  B = diag(2),
  U = matrix("u", nrow = 2, ncol = 1),
  Q = matrix(list("T", 0, 0, "W"), 2, 2),
  Z = diag(2),
  A = matrix(c(0, 0), nrow = 2, ncol = 1),
  R = matrix(list("T", 0, 0, "W"), 2, 2)
)

## fit model
size_both_2 <- MARSS(t(log(yy)), model = mod_list)
```


---

# Model forms

.blue-text[
## One common state
]

$$\begin{bmatrix} y_\text{Tengu} \\ y_\text{WDFW} \end{bmatrix}_t = \begin{bmatrix} 1 \\ 1 \end{bmatrix}
x_t + \begin{bmatrix} a_\text{Tengu} \\ a_\text{WDFW} \end{bmatrix} + \begin{bmatrix} v_\text{Tengu} \\
v_\text{WDFW} \end{bmatrix}_t$$

$$x_t = x_{t-1} + u + w_t$$

.blue-text[
## Two unique states
]

$$\begin{bmatrix} y_\text{Tengu} \\ y_\text{WDFW} \end{bmatrix}_t = \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix}
\begin{bmatrix} x_\text{Tengu} \\ x_\text{WDFW} \end{bmatrix}_t + \begin{bmatrix} a_\text{Tengu} \\ a_\text{WDFW} \end{bmatrix} + \begin{bmatrix} v_\text{Tengu} \\ v_\text{WDFW} \end{bmatrix}_t$$

$$\begin{bmatrix} x_\text{Tengu} \\ x_\text{WDFW} \end{bmatrix}_t = \begin{bmatrix} x_\text{Tengu} \\ x_\text{WDFW} \end{bmatrix}_{t-1} + \begin{bmatrix} u_\text{Tengu} \\ u_\text{WDFW} \end{bmatrix} + \begin{bmatrix} w_\text{Tengu} \\ w_\text{WDFW} \end{bmatrix}_t$$

---

class: middle, center

# The data support .green-text[two unique states] with a .green-text[common decline] of ~0.4% per year

---

# Model fits and data

```{r plot_ssm_one_size, fig.dim = c(8, 5), echo = FALSE}
par(mai = c(1.1, 1.3, 0.3, 0.3), cex.axis = 1.3, cex.lab = 1.6)
matplot(yy, type = "l", lty = "solid", lwd = 1, col = c("dodgerblue", "indianred"), las = 1,
        ylab = "Mean size (kg)", xlab = "Year", xaxt = "n")
matlines(t(exp(size_both_2$states)), type = "l", lty = "dashed", lwd = 2, col = c("dodgerblue", "indianred"))
text(17, 4, "Tengu", pos = 1, cex = 1.5, col = "dodgerblue")
text(17, 8.2, "WDFW", pos = 3, cex = 1.5, col = "indianred")
axis(1, at = seq(1, 41, 10), labels = seq(1970, 2010, 10))
```

---

# Conclusions

.green-text[
## Tengu anglers and WDFW assessments seem to be sampling different groups of fish

## Both data sources showed evidence for declining abundance and size over time

## The Tengu data are an amazing source of information
]

.futnote.purple-text[`r icon::fa("envelope")` scheuerl@uw.edu]

.citation.blue-text[`r icon::fa("twitter")` @mark_scheuerell]

