---
title: "Shifts in the size of Chinook salmon from a culturally important recreational fishery"
subtitle: "Supplementary material to accompany Quinn et al."
author: "Mark Scheuerell"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    highlight: haddock
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")

library(dplyr)
library(tidyr)
library(MARSS)

## plot hook to increase margins
local({
  hook_old <- knitr::knit_hooks$get("plot")  # save the old hook
  knitr::knit_hooks$set(plot = function(x, options) {
    x <- c("\\vspace{0.5in}", hook_old(x, options))
  })
})

## better captions
library("captioner")
## set default caption delimter
fig_cap <- captioner(suffix = ".", style = "b", style_prefix = TRUE)
```


# Data

The original Tengu Derby data were provided to me by Tom Quinn on 16 June 2020 in the form of an MS Excel file titled `Tengu_derby_leaders through 2019 derby.xls`. I exported one worksheet of interest (`data in kg`) as `~/data/tengu_derby_data.csv`. The same Excel file also included a worksheet with information from WDFW on the number of natural- and hatchery-origin Chinook, and the mean mass of Chinook (`Losee Chiook data`), which I exported as `~/data/wdfw_data.csv`

```{r import_data}
## set data dir
datadir <- here::here("data")
## import raw Tengu data
tengu_data <- readr::read_csv(file.path(datadir, "tengu_derby_data.csv"))
## import raw WDFW data
wdfw_data <- readr::read_csv(file.path(datadir, "wdfw_data.csv"))
```

# Changes in fish size over time

The data set contains three different indicators of fish size over time:

1) the total number of fish over 10 pounds (~4.55 kg);

2) the total number of fish over 5 pounds (~2.27 kg); and

3) the masses (kg) of the 5 largest fish.

Clearly (1) and (2) will be correlated, as (1) is a subset of (2). Furthermore, the probability of catching a fish greater than 5 or 10 pounds clearly increases as the total number of fish caught also increases. Thus, we calculated the proportion of fish caught in a given year that were greater than the two size thresholds.

## Proportions of large fish

```{r prop_of_fish}
## proportion of fish >5 lbs & >10 lbs
size_props <- tengu_data %>%
  mutate(
    ## proportion of fish >5 lbs
    p5 = n_over_5 / total_catch,
    ## proportion of fish >10 lbs
    p10 = n_over_10 / total_catch) %>%
  ## select cols of interest
  select(year, p5, p10) %>%
  ## remove years prior to 1950 with all NA's
  filter(!(is.na(p5) & is.na(p10)))
```

Here are plots of the two proportional size metrics over time.

```{r ts_plot_props, fig.dim = c(6, 4), echo = FALSE}
## time series plot of fish size metrics
par(mai = c(0.9, 0.9, 0.1, 0.1))
matplot(size_props[,1], size_props[,(2:3)],
        las = 1, lwd = 2, type = "l", lty = "solid",
        col = c("dodgerblue", "darkblue"),
        xlab = "Year", ylab = "Proportion exceeding")
text(1955, 0.65, "5 lbs", col = "dodgerblue")
text(1955, 0.15, "10 lbs", col = "darkblue")
```

\setlength{\leftskip}{0.5in}
\setlength{\rightskip}{0.5in}

\small

`r fig_cap(1, caption = "Time series of the proportion of fish caught in the Tengu Derby that exceeded 5 pounds (light blue) and 10 pounds (dark blue).", display = "full")` 

\normalsize

\setlength{\leftskip}{0in}
\setlength{\rightskip}{0in}

\vspace{0.25in}

There appears to be an overall decline in fish size from 1950 until the early 1980s, when fish sizes increased rapidly before declining again until present.

## Mean of 5 largest fish

```{r mean_of_big_fish}
## get mean of top-5; code as NA if <5 fish were weighed
tengu_sizes <- tengu_data %>%
  #filter(year >= 1970 & year <= 2014) %>%
  select(starts_with(c("year", "size"))) %>%
  rowwise(year) %>% 
  summarise(tengu_mean_kg = mean(c_across(everything())))
```

```{r ts_plot_means, fig.dim = c(6, 4), echo = FALSE}
## time series plot of fish size metrics
par(mai = c(0.9, 0.9, 0.1, 0.1))
plot(tengu_sizes$year, tengu_sizes$tengu_mean_kg,
     las = 1, xaxt = "n",
     type = "o", lty = "solid", lwd = 2,
     pch = 16, col = "dodgerblue",
     xlab = "Year", ylab = "Mean mass (kg)")
axis(1, at = seq(1950, 2020, 10))
```

\setlength{\leftskip}{0.5in}
\setlength{\rightskip}{0.5in}

\small

`r fig_cap(2, caption = "Time series of the mean mass of the five largest fish caught in the Tengu Derby.", display = "full")` 

\normalsize

\setlength{\leftskip}{0in}
\setlength{\rightskip}{0in}

\vspace{0.25in}

# Models for Tengu Derby fish mass

## Unbiased random walk

```{r fit_tengu_urw}
## model defn
mod_list <- list(
  B = matrix(1),
  U = matrix(0),
  Q = matrix("q"),
  Z = matrix(1),
  R = matrix("r")
)

## data for {MARSS}
yy <- matrix(tengu_sizes$tengu_mean_kg, nrow = 1) %>%
  log()

## fit unbiased RW
size_tengu_urw <- MARSS(yy, model = mod_list)
```


## Biased random walk

```{r fit_tengu_brw}
## update model defn
mod_list$U <- matrix("u")

## fit biased RW
size_tengu_brw <- MARSS(yy, model = mod_list)
```

## Compare models

```{r compare_models}
## AICc for unbiased RW
size_tengu_urw$AICc
## AICc for biased RW
size_tengu_brw$AICc
```

## Plot model fit

```{r ts_plot_fits, fig.dim = c(6, 4), echo = FALSE}
## calculate approx 95% CI's
upCI <- exp(size_tengu_urw$states + 2*size_tengu_urw$states.se)
loCI <- exp(size_tengu_urw$states - 2*size_tengu_urw$states.se)

## time series plot of fish size metrics
par(mai = c(0.9, 0.9, 0.1, 0.1))
plot(tengu_sizes$year, tengu_sizes$tengu_mean_kg,
     las = 1, xaxt = "n",
     type = "o", lty = "solid", lwd = 2,
     pch = 16, col = "dodgerblue",
     xlab = "Year", ylab = "Mean mass (kg)")
lines(tengu_sizes$year,
      t(upCI),
      col = "gray", lty = "solid", lwd = 2,)
lines(tengu_sizes$year,
      t(exp(size_tengu_urw$states)),
      col = "black", lty = "solid", lwd = 2,)
lines(tengu_sizes$year,
      t(loCI),
      col = "gray", lty = "solid", lwd = 2,)
axis(1, at = seq(1950, 2020, 10))
```

\setlength{\leftskip}{0.5in}
\setlength{\rightskip}{0.5in}

\small

`r fig_cap(3, caption = "Time series of the mean mass of the five largest fish caught in the Tengu Derby (solid) and the model fitted values (dashed).", display = "full")` 

\normalsize

\setlength{\leftskip}{0in}
\setlength{\rightskip}{0in}

\vspace{0.25in}



# Comparisons between the Tengu Derby & WDFW

## Combined data sets

The time series from WDFW begins in 1970 and runs through 2019, so we restricted our analysis to the 45 years from 1970-2019. Again, we fit models to the log-transformed size data.

```{r both_sizes}
## get corresponding size data from WDFW
wdfw_sizes <- wdfw_data %>%
  ## drop first 2 years with NA's for size
  drop_na(size) %>%
  ## keep year & size
  select(year, size) %>%
  ## rename size
  rename(wdfw_mean_kg = size)

## combine data sets for {MARSS}
y2 <- tengu_sizes %>%
  left_join(wdfw_sizes, by = "year") %>%
  ungroup() %>%
  ## drop non-overlapping years
  filter(year >= 1970 & year <= 2019) %>%
  ## drop year
  select(-year) %>%
  ## log-transform
  log() %>%
  t()
```

## Plot data

```{r plot_both_sizes, fig.dim = c(6, 4), echo = FALSE}
par(mai = c(0.9, 0.9, 0.1, 0.1))
matplot(t(exp(y2)), type = "o", lty = "solid", lwd = 2, pch = 16,
        las = 1,
        col = c("dodgerblue", "indianred"),
        ylab = "Mean mass (kg)", xlab = "Year", xaxt = "n")
axis(1, at = seq(1, 41, 10), labels = seq(1970, 2010, 10))
```


## One-state models

### Unbiased random walk

```{r ssm_one_size}
## model defn
mod_list <- list(
  B = matrix(1),
  U = matrix(0),
  Q = matrix("q"),
  Z = matrix(1, nrow = 2, ncol = 1),
  A = matrix(c("T", "W"), nrow = 2, ncol = 1),
  R = matrix(list("r", 0, 0, "r"), 2, 2)
)

## fit unbiased RW (Eqns 5 & 6)
size_both_1u <- MARSS(y2, model = mod_list)
```

### Biased random walk

```{r}
## update model list
mod_list$U <- matrix("u")

## fit biased RW (Eqns ? & ?)
size_both_1b <- MARSS(y2, model = mod_list)
```


## Two-state models

### Unbiased random walk

```{r ssm_two_size}
## model defn
mod_list <- list(
  B = diag(2),
  U = matrix(0, nrow = 2, ncol = 1),
  Q = matrix(list("T", 0, 0, "W"), 2, 2),
  # Q = matrix(list("q", 0, 0, "q"), 2, 2),
  Z = diag(2),
  # A = matrix(c("T", "W"), nrow = 2, ncol = 1),
  A = matrix(c(0, 0), nrow = 2, ncol = 1),
  # R = matrix(list("r", 0, 0, "r"), 2, 2)
  R = matrix(list("T", 0, 0, "W"), 2, 2)
)

## fit biased RW (Eqns ? & ?)
size_both_2u <- MARSS(y2, model = mod_list)
```

### Random walk with unique biases

```{r}
## update model defn
mod_list$U <- matrix(c("T", "W"), nrow = 2, ncol = 1)

## fit biased RW
size_both_2b_unique <- MARSS(y2, model = mod_list)
```

### Random walk with shared bias

```{r}
## update model defn
mod_list$U <- matrix(c("u", "u"), nrow = 2, ncol = 1)

## fit biased RW
size_both_2b_shared <- MARSS(y2, model = mod_list)
```

## Compare models

```{r compare_models_both}
## AICc for unbiased RW
size_both_2u$AICc
## AICc for biased RW with unique bias
size_both_2b_unique$AICc
## AICc for biased RW with shared bias
size_both_2b_shared$AICc
```

The unbiased random walk has an AICc score that is slightly greater than the biased random walk, but it's more simple and should be favored here.

## Plot model fits

```{r plot_ssm_two_sizes, fig.dim = c(6, 4), echo = FALSE}
par(mai = c(0.9, 0.9, 0.1, 0.1))
matplot(t(exp(y2)), type = "o", lty = "solid", lwd = 2, pch = 16,
        las = 1,
        col = c("dodgerblue", "indianred"),
        ylab = "Mean mass (kg)", xlab = "Year", xaxt = "n")
matlines(t(exp(size_both_2b_shared$states)),
         type = "l", lty = "dashed", lwd = 1, col = c("dodgerblue", "indianred"))
# matlines(t(exp(size_both_2u$states)),
#          type = "l", lty = "dashed", lwd = 1, col = c("dodgerblue", "indianred"))
axis(1, at = seq(1, 41, 10), labels = seq(1970, 2010, 10))
```

\setlength{\leftskip}{0.5in}
\setlength{\rightskip}{0.5in}

\small

`r fig_cap(6, caption = "Time series of observed fish size from the Tengu derby (blue) and WDFW surveys (red), including fits from the multivariate random walk model for both series (dashed lines).", display = "full")` 

\normalsize

\setlength{\leftskip}{0in}
\setlength{\rightskip}{0in}

\vspace{0.25in}

### Summary of size comparison

The model with one common state has an AICc value of `r round(size_both_1u$AICc, 1)` and the model with two unique states has an AICc value of `r round(size_both_2u$AICc, 1)`, which indicates rather strong support for two unique temporal patterns in the data.


# Additional figures for paper

## Plot fitted values for one-state model

```{r plot_ssm_one_size, fig.dim = c(6, 4), echo = FALSE, eval = FALSE}
par(mai = c(0.9, 0.9, 0.1, 0.1))
matplot(t(exp(y2)), type = "o", lty = "solid", lwd = 2, pch = 16,
        col = c("dodgerblue", "indianred"),
        ylab = "Mean size (kg)", xlab = "Year", xaxt = "n")
matlines(t(exp(size_both_1u$states)), type = "l", lty = "dashed", lwd = 1, col = c("black"))
axis(1, at = seq(1, 41, 10), labels = seq(1970, 2010, 10))
```

\setlength{\leftskip}{0.5in}
\setlength{\rightskip}{0.5in}

\small

`r fig_cap(5, caption = "Time series of observed fish size from the Tengu derby (blue) and WDFW surveys (red), including fits from the multivariate random walk model (Eqns 6 & 7; dashed).", display = "full")` 

\normalsize

\setlength{\leftskip}{0in}
\setlength{\rightskip}{0in}

\vspace{0.25in}

## Other size metrics

## Proportions of large fish

```{r prop_of_big_fish}
## proportion of fish >5 lbs & >10 lbs
big_per_angler <- tengu_data %>%
  mutate(
    ## proportion of fish >5 lbs
    b5 = n_over_5 / members,
    ## proportion of fish >10 lbs
    b10 = n_over_10 / members) %>%
  ## select cols of interest
  select(year, b5, b10)
```

Here are some additional plots for the supplemental material.

```{r plot_size_metrics, fig.dim = c(6, 8), message = FALSE}
tengu_max <- tengu_data %>%
  #filter(year >= 1970 & year <= 2019) %>%
  select(starts_with(c("year", "size"))) %>%
  rowwise(year) %>% 
  summarise(tengu_max_kg = max(c_across(everything())))

## plot other size metrics
par(mfrow = c(2,1),
    mai = c(0.9, 0.9, 0.1, 0.1))

## largest fish
plot(tengu_max$year, tengu_max$tengu_max_kg,
     las = 1, type = "o", pch = 16, lwd = 2,
     col = "purple", xlim = c(1945, 2020),
     xlab = "Year",
     ylab = "Maximum mass (kg)")

## proportion exceeding
matplot(big_per_angler[,1], big_per_angler[,(2:3)],
        las = 1, pch = 16, xlim = c(1945, 2020),
        lwd = 2, type = "o", lty = "solid",
        col = c("dodgerblue", "darkblue"),
        xlab = "Year", ylab = "Number of large fish per angler")
text(1955, 0.72, "5 lbs", col = "dodgerblue")
text(1955, 0.12, "10 lbs", col = "darkblue")
```


