---
title: "Shifts in the size of fish from a culturally important recreational fishery"
author: "Mark Scheuerell"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    highlight: haddock
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")

library(dplyr)
library(MARSS)

## plot hook to increase margins
local({
  hook_old <- knitr::knit_hooks$get("plot")  # save the old hook
  knitr::knit_hooks$set(plot = function(x, options) {
    x <- c("\\vspace{0.5in}", hook_old(x, options))
  })
})

## better captions
library("captioner")
## set default caption delimter
fig_cap <- captioner(suffix = ".", style = "b", style_prefix = TRUE)
```


# Data

The original Tengu Derby data were provided to me by Tom Quinn on 16 June 2020 in the form of an MS Excel file titled `Tengu_derby_leaders through 2019 derby.xls`. I exported one worksheet of interest (`data in kg`) as `~/data/tengu_derby_data.csv`. The same Excel file also included a worksheet with information from WDFW on the number of natural- and hatchery-origin Chinook, and the mean mass of Chinook (`Losee Chiook data`), which I exported as `~/data/wdfw_data.csv`

```{r import_data}
## set data dir
datadir <- here::here("data")
## import raw Tengu data
tengu_data <- readr::read_csv(file.path(datadir, "tengu_derby_data.csv"))
## import raw WDFW data
wdfw_data <- readr::read_csv(file.path(datadir, "wdfw_data.csv"))
```

# Changes in fish size over time

The data set contains three different indicators of fish size over time:

1) the total number of fish over 10 pounds (~4.55 kg);

2) the total number of fish over 5 pounds (~2.27 kg); and

3) the masses (kg) of the 5 largest fish.

Clearly (1) and (2) will be correlated, as (1) is a subset of (2). Furthermore, the probability of catching a fish greater than 5 or 10 pounds clearly increases as the total number of fish caught also increases. Thus, I modeled the proportion of fish caught in a given year that were greater than the 2 size thresholds.

```{r prop_of_fish}
## proportion of fish >10 lbs
p10 <- tengu_data$n_over_10 / tengu_data$total_catch
## screen for p = 0 & change to p = 0.001
p10[p10 == 0] <- 0.001

## proportion of fish >5 lbs
p5 <- tengu_data$n_over_5 / tengu_data$total_catch
## screen for p = 1 & change to p = 0.999
p5[p5 == 1] <- 0.999

## combine proportional data
fish_sizes <- cbind("Prop. over 10 lbs" = p10, "Prop. over 5 lbs" = p5) %>%
  ts(start = min(tengu_data$year))
```


Here are plots of the two size metrics over time.

```{r ts_plot_sizes, fig.dim = c(6, 6), echo = FALSE}
## time series plot of fish size metrics
par(mai = c(0.9, 0.9, 0.1, 0.1))
plot.ts(fish_sizes, lwd = 2, col = "dodgerblue",
        main = "", yax.flip = TRUE,
        ylab = c("", "", ""), xlab = "Year")
```

\setlength{\leftskip}{0.5in}
\setlength{\rightskip}{0.5in}

\small

`r fig_cap(3, caption = "Time series of the proportion of fish over 10 pounds (top) and over 5 pounds (bottom).", display = "full")` 

\normalsize

\setlength{\leftskip}{0in}
\setlength{\rightskip}{0in}

\vspace{0.25in}

There appears to be an overall decline in fish size from the mid 1940s until the early 1980s, when fish sizes increased rapidly before declining again until present.

## Check annual size data
size_check <- tengu_data %>%
  select(starts_with("size")) %>%
  apply(1, is.na) %>%
  t()

## Fit biased random walk

Just as I did with the CPUE data, I fit both biased and unbiased forms of random walk models to the fish size data, with the response being the logit-transformed proportions of fish over 10 and 5 pounds.

```{r model_setup_2}
## response for fish >10 lbs
l_size_10 <- matrix(qlogis(p10), nrow = 1)

## response for fish >5 lbs
l_size_5 <- matrix(qlogis(p5), nrow = 1)

## model setup
mod_list <- list(
  B = matrix(1),
  U = matrix("u"),
  Q = matrix("q"),
  Z = matrix(1),
  A = matrix(0),
  R = matrix("r")
)
```

```{r biased_RW_size}
## over 10
## fit model with bias (Eqn 1)
size_brw_10 <- MARSS(l_size_10, model = mod_list)

## 95% CI for bias
size_brw_10 <- MARSSparamCIs(size_brw_10)


## over 5
## fit model with bias (Eqn 1)
size_brw_5 <- MARSS(l_size_5, model = mod_list)

## 95% CI for bias
size_brw_5 <- MARSSparamCIs(size_brw_5)
```

## Fit unbiased random walk

```{r unbiased_RW_size}
## set bias to 0
mod_list$U <- matrix(0)

## over 10
## fit model without bias (Eqn 2)
size_rw_10 <- MARSS(l_size_10, model = mod_list)

## over 5
## fit model without bias (Eqn 2)
size_rw_5 <- MARSS(l_size_5, model = mod_list)
```

\vspace{0.25in}

For the model based upon the proportion of fish over 10 pounds, the 95% confidence interval for the bias term $(u)$ is (`r round(c(size_brw_10$par.lowCI$U, size_brw_10$par.upCI$U), 3)`), and the model with a bias term has an AIC value that is only ~`r round(size_rw_10$AICc - size_brw_10$AICc, 1)` units lower than the model without a bias term, suggesting there is essentially no data support for a systematic downward trend in the log-transformed size over time. This is not much of a surprise, however, given the apparent temporal patterns in the data.

For the model based upon the proportion of fish over 5 pounds, the 95% confidence interval for the bias term $(u)$ is (`r round(c(size_brw_5$par.lowCI$U, size_brw_5$par.upCI$U), 3)`), and the model with a bias term has an AIC value that is only ~`r round(size_rw_5$AICc - size_brw_5$AICc, 1)` units lower than the model without a bias term, suggesting there is essentially no data support for a systematic downward trend in the log-transformed size over time. 

## Model fit to size data

```{r plot_size_10_rw, fig.dim = c(6, 4), echo = FALSE}
## time series plot of size & model fit
par(mai = c(0.9, 0.9, 0.1, 0.1))
plot.ts(ts(t(plogis(l_size_10)), start = min(tengu_data$year)), lwd = 2, col = "dodgerblue",
        las = 1, ylab = "Prop. over 10 lbs", xlab = "Year")
lines(ts(t(plogis(size_rw_10$states)), start = min(tengu_data$year)))
```

\setlength{\leftskip}{0.5in}
\setlength{\rightskip}{0.5in}

\small

`r fig_cap(4, caption = "Time series of observed size (blue) and the fit for the random walk model (Eqn 3; black).", display = "full")` 

\normalsize

\setlength{\leftskip}{0in}
\setlength{\rightskip}{0in}

\vspace{0.25in}

```{r plot_size_5_rw, fig.dim = c(6, 4), echo = FALSE}
## time series plot of size & model fit
par(mai = c(0.9, 0.9, 0.1, 0.1))
plot.ts(ts(t(plogis(l_size_5)), start = min(tengu_data$year)), lwd = 2, col = "dodgerblue",
        las = 1, ylab = "Prop. over 5 lbs", xlab = "Year")
lines(ts(t(plogis(size_rw_5$states)), start = min(tengu_data$year)))
```

\setlength{\leftskip}{0.5in}
\setlength{\rightskip}{0.5in}

\small

`r fig_cap(5, caption = "Time series of observed size (blue) and the fit for the random walk model (Eqn 3; black).", display = "full")` 

\normalsize

\setlength{\leftskip}{0in}
\setlength{\rightskip}{0in}

\vspace{0.25in}

# Comparisons between the Tengu Derby & WDFW

Although the two data sources come from different times, places, and gear types, they both contain information on the temporal changes in size and CPUE over time. I investigated whether or not the temporal trends in the two data sources track one another (i.e., are representative of one "state of nature"). To do so, I used a multivariate state-space model of the general form

\begin{gather}
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
\end{gather}

For both forms of the model, $\mathbf{y}_t$ is a $[2 \times 1]$ vector of the observed data from both sources, $\mathbf{a}$ is a $[2 \times 1]$ vector of offsets (intercepts), and $\mathbf{v}_t \sim \text{MVN}(\mathbf{0}, \mathbf{R})$. For both models, I assumed that the observation errors at time $t$ $(\mathbf{v}_t)$ are independent and differently distributed, such that 

$$
\mathbf{R} =
\begin{bmatrix}
r_1 & 0 \\
0 & r_2
\end{bmatrix}
$$

## Fish size

The time series from WDFW begins in 1970 and runs through 2015, but the Tengy Derby data is missing size information for 2015, so I restricted my analysis to the 45 years from 1970-2014. Again I fit models to the log-transformed size data.

### One pattern over time

For the model with only one state of nature, $\mathbf{Z}$ is a $[2 \times 1]$ vector of 1's, $\mathbf{x}_t$ is a $[1 \times 1]$ scalar of the true state, and $\mathbf{w}_t \sim \text{N}(0, q)$, such that

\begin{gather}
\begin{bmatrix}
y_\text{Tengu} \\
y_\text{WDFW}
\end{bmatrix}_t
 = 
\begin{bmatrix}
1 \\
1
\end{bmatrix}
x_t + 
\begin{bmatrix}
a_\text{Tengu} \\
a_\text{WDFW}
\end{bmatrix}
+ 
\begin{bmatrix}
v_\text{Tengu} \\
v_\text{WDFW}
\end{bmatrix}_t
\\
x_t = x_{t-1} + w_t
\end{gather}

```{r mean_of_top_2}
## mean mass of 2 largest fish
mean_of_top_2 <- apply(tengu_data[,c(9:10)], 1, mean, na.rm = TRUE)
mean_of_top_2[is.nan(mean_of_top_2)] <- NA

## get ts of Tengu catch
catch <- tengu_data$total_catch[tengu_data$year >= 1970 & tengu_data$year <= 2014]

## select common data
yy <- cbind(tengu = mean_of_top_2[tengu_data$year >= 1970 & tengu_data$year <= 2014],
            wdfw = wdfw_data$size[wdfw_data$year >= 1970 & wdfw_data$year <= 2014])
```

```{r ssm_one_size}
## model defn for Eqns 6 & 7
mod_list <- list(
  B = matrix(1),
  U = matrix(0),
  Q = matrix("q"),
  Z = matrix(1, nrow = 2, ncol = 1),
  A = matrix(c("T", "W"), nrow = 2, ncol = 1),
  # R = matrix(list("T", 0, 0, "W"), 2, 2)
  R = matrix(list("r", 0, 0, "r"), 2, 2)
)

## fit Eqns 6 & 7
size_both_1 <- MARSS(t(log(yy)), model = mod_list)
```

```{r plot_ssm_one_size, fig.dim = c(6, 4), echo = FALSE}
par(mai = c(0.9, 0.9, 0.1, 0.1))
matplot(yy, type = "l", lty = "solid", lwd = 2, col = c("dodgerblue", "indianred"),
        ylab = "Mean size (kg)", xlab = "Year", xaxt = "n")
matlines(t(exp(size_both_1$states)), type = "l", lty = "dashed", lwd = 1, col = c("black"))
axis(1, at = seq(1, 41, 10), labels = seq(1970, 2010, 10))
```

\setlength{\leftskip}{0.5in}
\setlength{\rightskip}{0.5in}

\small

`r fig_cap(5, caption = "Time series of observed fish size from the Tengu derby (blue) and WDFW surveys (red), including fits from the multivariate random walk model (Eqns 6 & 7; dashed).", display = "full")` 

\normalsize

\setlength{\leftskip}{0in}
\setlength{\rightskip}{0in}

\vspace{0.25in}


### Two patterns over time

For the model with two different states of nature, $\mathbf{Z}$ is a $[2 \times 2]$ identity matrix, $\mathbf{x}_t$ is a $[2 \times 1]$ vector of the true states, and $\mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})$, such that

\begin{gather}
\begin{bmatrix}
y_\text{Tengu} \\
y_\text{WDFW}
\end{bmatrix}_t
 = 
\begin{bmatrix}
1 & 0 \\
0 & 1
\end{bmatrix}
\begin{bmatrix}
x_\text{Tengu} \\
x_\text{WDFW}
\end{bmatrix}_t
+ 
\begin{bmatrix}
a_\text{Tengu} \\
a_\text{WDFW}
\end{bmatrix}
+ 
\begin{bmatrix}
v_\text{Tengu} \\
v_\text{WDFW}
\end{bmatrix}_t
\\
\begin{bmatrix}
x_\text{Tengu} \\
x_\text{WDFW}
\end{bmatrix}_t
= 
\begin{bmatrix}
x_\text{Tengu} \\
x_\text{WDFW}
\end{bmatrix}_{t-1}
+ 
\begin{bmatrix}
w_\text{Tengu} \\
w_\text{WDFW}
\end{bmatrix}_t
\end{gather}

```{r ssm_two_size}
## model defn for Eqns 8 & 9
mod_list <- list(
  B = diag(2),
  U = matrix("u", nrow = 2, ncol = 1),
  # Q = matrix(list("T", 0, 0, "W"), 2, 2),
  Q = matrix(list("q", 0, 0, "q"), 2, 2),
  Z = diag(2),
  A = matrix(c(0, 0), nrow = 2, ncol = 1),
  # R = matrix(list("T", 0, 0, "W"), 2, 2)
  R = matrix(list("r", 0, 0, "r"), 2, 2)
)

## fit Eqns 8 & 9
size_both_2 <- MARSS(t(log(yy)), model = mod_list)
```

```{r plot_ssm_two_size, fig.dim = c(6, 4), echo = FALSE}
par(mai = c(0.9, 0.9, 0.1, 0.1))
matplot(yy, type = "l", lty = "solid", lwd = 2, col = c("dodgerblue", "indianred"),
        ylab = "Mean size (kg)", xlab = "Year", xaxt = "n")
matlines(t(exp(size_both_2$states)), type = "l", lty = "dashed", lwd = 1, col = c("dodgerblue", "indianred"))
axis(1, at = seq(1, 41, 10), labels = seq(1970, 2010, 10))
```

\setlength{\leftskip}{0.5in}
\setlength{\rightskip}{0.5in}

\small

`r fig_cap(6, caption = "Time series of observed fish size from the Tengu derby (blue) and WDFW surveys (red), including fits from the multivariate random walk model (Eqns 8 & 9; dashed).", display = "full")` 

\normalsize

\setlength{\leftskip}{0in}
\setlength{\rightskip}{0in}

\vspace{0.25in}

### Summary of size comparison

The model with one common state has an AICc value of `r round(size_both_1$AICc, 1)` and the model with two unique states has an AICc value of `r round(size_both_2$AICc, 1)`, which indicates rather modest support for two unique temporal patterns in the data.




